---
title: "seiqhrf_model"
output: html_document
always_allow_html: true
---

```{r setup, include=FALSE, eval=TRUE}
version_date <- lubridate::ymd(Sys.Date())
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, eval=TRUE,
                      tidy.opts=list(width.cutoff=60),
                      tidy=TRUE)
```

## Libraries

```{r, echo=TRUE, eval=TRUE, message=FALSE}
library(tidyverse)
library(magrittr)
library(lubridate)
library(stringr)
library(tibble)
library(broom)
library(ggplot2)
library(gt)
library(knitr)
library(devtools)
library(DiagrammeR)
library(parallel)
library(foreach)
library(tictoc)
suppressMessages(library(EpiModel))
library(incidence)
library(earlyR)
```


## The SEIQHRF model

| Compartment | Functional definition                                                            |
|-------------|----------------------------------------------------------------------------------|
| S           | Susceptible individuals                                                          |
| E           | Exposed **and** infected, not yet symptomatic but potentially infectious         |
| I           | Infected, symptomatic **and** infectious                                         |
| Q           | Infectious, but (self-)isolated                                                  |
| H           | Requiring hospitalisation (would normally be hospitalised if capacity available) |
| R           | Recovered, immune from further infection                                         |
| F           | Case fatality (death due to COVID-19, not other causes)                          |

### Transition diagram

```{r, echo=FALSE, eval=TRUE, message=FALSE, layout="l-page"}
grViz("
digraph SEIQHRF {
  # a 'graph' statement
  graph [overlap = false, fontsize = 10] #, rankdir = LR]
  # several 'node' statements
  node [shape = box,
        fontname = Helvetica]
  S[label='S=Susceptible'];
  E[label='E=Exposed and infected,\nasymptomatic,\npotentially infectious'];
  I[label='I=Infected and infectious'];
  Q[label='Q=(Self-)isolated\n(infectious)'];
  H[label='H=Requires\nhospitalisation'];
  R[label='R=Recovered/immune'];
  F[label='F=Case fatality']
  # several 'edge' statements
  S->E[label='a']
  I->S[style='dashed', label='x']
  E->I[label='b']
  E->S[style='dashed', label='y']
  I->Q[label='c']
  Q->S[style='dashed', label='z']
  I->R[label='d']
  I->H[label='e']
  H->F[label='f']
  H->R[label='g']
  Q->R[label='h']
  Q->H[label='i']
}
")
```

These functions extend the `EpiModel` code to define a SEIQHRF model.


```{r, echo=TRUE, eval=TRUE, message=FALSE}
source('model/_icm.mod.init.seiqhrf.R')
source('model/_icm.mod.status.seiqhrf.R')
source('model/_icm.mod.vital.seiqhrf.R')
source('model/_icm.control.seiqhrf.R')
source('model/_icm.utils.seiqhrf.R')
source('model/_icm.saveout.seiqhrf.R')
source('model/_icm.icm.seiqhrf.R')
```

## Parameters

The `simulate()` function provides baseline defaults for the simulation parameters. The default values result in a model that behaves in a way that mimics dynamic models. Public health interventions can then be investigated by changing just those parameters which relate to or represent the intervention. 

The main parameters of interest are as follows:

```{r, echo=FALSE, eval=TRUE, message=FALSE, layout="l-page"}
param_docs <- tribble(~DiagramRef, ~Parameter, ~Default, ~Explanation,
 "", "type", "SEIQHRF", "Type of model: SI, SIR, SIS, SEIR, SEIQHR and SEIQHRF available, but only SEIQHRF is likely to work in the current version of the code.",
  "", "nsteps", "366", "Number of days for simulation. Note that day 1 is for initialisation, day 2 is the first day of the simulation, hence default of 366 for 1 year.",
  "", "nsims", "10", "Number of simulations to run and then average.",
  "", "ncores", "10", "Number of CPU cores to use for parallel execution.",
  "b", "prog.rand", "FALSE", "Method for progression from E compartment to I. If TRUE, random binomial draws at `prog.rate`, if FALSE, random draws from a Weibull distribution, with parameters `prog.dist.scale` and  `prog.dist.shape`",
  "d,g,h", "rec.rand", "FALSE", "Method for recovery transition from I, Q or H to R. If TRUE, random binomial draws at `prog.rate`, if FALSE, random draws from a Weibull distribution, with parameters `rec.dist.scale` and  `rec.dist.shape`",
  "f", "fat.rand", "FALSE", "Method for case fatality transition from H to F. If TRUE, random binomial draws at `fat.rate.base`, if FALSE, random sample with a sample fraction also given by `fat.rate.base`. However, if the current number of patients in the H (needs hospitalisation) compartment is above a hospital capacity level specified by `hosp.cap`, then the fatality rate is the mean of the base fatality rate weighted by the hospital capacity, plus a higher rate, specified by `fat.rate.overcap`, weighted by the balance of those requiring hospitalisation (but who can't be accommodated). By setting `fat.rate.overcap` higher, the effect of exceeding the capacity of the health care system can be simulated. There is also a coefficient `fat.tcoeff` for the fatality rates that increases them as a linear function of the number of days the individual has been in the H compartment. Use of the co-efficient better approximates the trapezoid survival time distribution typical of ICU patients.",
  "c", "quar.rand", "FALSE", "Method for self-isolation transition from I to Q. If TRUE, random binomial draws at `quar.rate`, if FALSE, random sample with a sample fraction also given by `quar.rate`.",
  "e,i", "hosp.rand", "FALSE", "Method for transition from I or Q to H -- that is, from infectious or from self-isolated to requiring hospitalisation. If TRUE, random binomial draws at `hosp.rate`, if FALSE, random sample with a sample fraction also given by `hosp.rate.",
  "e,i", "disch.rand", "FALSE", "Method for transition from H to R -- that is, from requiring hospitalisation to recovered. If TRUE, random binomial draws at `disch.rate`, if FALSE, random sample with a sample fraction also given by `disch.rate`. Note that the only way out of the **H** compartment is recovery or death.  ",
  "", "infection.FUN", "infection.seiqhrf.icm", "No, being infected with SARS-CoV2 is **not** fun. Rather this is the the name of the function to implement infection processes. Use the default.",
 
  "", "departures.FUN", "departures.seiqhrf.icm", "Handles background demographics, specifically departures (deaths not due to the virus, and emigration). Use the default.",
  "", "arrivals.FUN", "arrivals.icm", "Handles background demographics, specifically arrivals (births and immigration). Uses the original EpiModel code currently. A replacement that implements modelling the arrival of infected individuals is under development -- but for now, all arrivals go into the **S** compartment.",
  "", "get_prev.FUN", "get_prev.seiqhrf.icm", "Utility function that collects prevalence and transition time data from each run and stores it away in the simulation result object. Use the default.",
 
  "", "s.num", "9997", "Initial number of **S* compartment individuals in the simulated population. An overall population of 10,000 is a good compromise. A set of models will still take several minutes or more to run, in parallel. ",
  "", "e.num", "0", "Initial number of **E** compartment individuals in the simulated population.",
  "", "i.num", "3", "Initial number of **I** compartment individuals in the simulated population.",
  "", "q.num", "0", "Initial number of **Q** compartment individuals in the simulated population.",
  "", "h.num", "0", "Initial number of **H** compartment individuals in the simulated population.",
  "", "r.num", "0", "Initial number of **R** compartment individuals in the simulated population.",
  "", "f.num", "0", "Initial number of **F** compartment individuals in the simulated population.",
  "x", "act.rate.i", "10", "The number of exposure events (_acts_) between infectious individuals in the **I** compartment and susceptible individuals in the **S** compartment, per day. It's stochastic, so the rate is an average, some individuals may have more or less. Note that not every exposure event results in infection - that is governed by the `inf.prob.i` parameters (see below). Reducing `act.rate.i` is equivalent to increasing social distancing by people in the **I** compartment.",
  "x", "inf.prob.i", "0.05", "Probability of passing on infection at each exposure event for interactions between infectious people in the **I** compartment and susceptibles in **S**. Reducing `inf.prob.i` is equivalent to increasing hygiene measures, such as not putting hands in eyes, nose or moth, use of hand sanitisers, wearing masks by the infected, and so on.",
  "y", "act.rate.e", "10", "The number of exposure events (_acts_) between infectious individuals in the **E** compartment and susceptible individuals in the **S** compartment, per day. Otherwise as for `act.rate.i`.",
  "y", "inf.prob.e", "0.02", "Probability of passing on infection at each exposure event for interactions between infectious people in the **E** compartment and susceptibles in **S**. Note the default is lower than for `inf.prob.i` reflecting the reduced infectivity of infected but asymptomatic people (the **E** compartment). Otherwise as for `inf.exp.i`.",
  "z", "act.rate.q", "2.5", "The number of exposure events (_acts_) between infectious individuals in the **Q** compartment (isolated, self or otherwise) and susceptible individuals in the **S** compartment, per day. Note the much lower rate than for the **I** and **E** compartments, reflecting the much greater degree of social isolation for someone in (self-)isolation. The exposure event rate is not zero for this group, just much less. Otherwise as for `act.rate.i`.",
  "z", "inf.prob.q", "0.02", "Probability of passing on infection at each exposure event for interactions between infectious people in the **Q** compartment and susceptibles in **S**. Note the default is lower than for `inf.prob.i` reflecting the greater care that self-isolated individuals will, on average, take regarding hygiene measures, such as wearing masks, to limit spread to others. Otherwise as for `inf.exp.i`.",
  "c", "quar.rate", "1/30", "Rate per day at which symptomatic (or tested positive), infected **I** compartment people enter self-isolation (**Q** compartment). Asymptomatic **E** compartment people can't enter self-isolation because they don't yet know they are infected. Default is a low rate reflecting low community awareness or compliance with self-isolation requirements or practices, but this can be tweaked when exploring scenarios.",
  "e,i", "hosp.rate", "1/100", "Rate **per day** at which symptomatic (or tested positive), infected **I** compartment people or self-isolated **Q** compartment people enter the state of requiring hospital care -- that is, become serious cases. A default rate of 1% per day with an average illness duration of about 10 days means a bit less than 10% of cases will require hospitalisation, which seems about right (but can be tweaked, of course).",
  "g", "disch.rate", "1/15", "Rate per day at which people needing hospitalisation recover.",
  "b", "prog.rate", "1/10", "Rate per day at which people who are infected but asymptomatic (**E** compartment) progress to becoming symptomatic (or test-positive), the **I** compartment. See `prog.rand` above for more details.",
  "b", "prog.dist.scale", "5", "Scale parameter for Weibull distribution for progression, see `prog.rand` for details.",
  "b", "prog.dist.shape", "1.5", "Shape parameter for Weibull distribution for progression, see `prog.rand` for details. Read up on the Weibull distribution before changing the default.",
                         
  "d", "rec.rate", "1/20", "Rate per day at which people who are infected and symptomatic (**I** compartment) recover, thus entering the **R** compartment. See `rec.rand` above for more details.",
  "d", "rec.dist.scale", "35", "Scale parameter for Weibull distribution for recovery, see `rec.rand` for details.",
  "d", "rec.dist.shape", "1.5", "Shape parameter for Weibull distribution for recovery, see `rec.rand` for details. Read up on the Weibull distribution before changing the default.",
  "f", "fat.rate.base", "1/50", "Baseline mortality rate per day for people needing hospitalisation (deaths due to the virus). See `fat.rand` for more details.",
  "f", "hosp.cap", "40", "Number of available hospital beds for the modelled population. See `fat.rand` for more details.",
  "f", "fat.rate.overcap", "1/25", "Mortality rate per day for people needing hospitalisation but who can't get into hospital due to the hospitals being full (see `hosp.cap` and `fat.rand`). The default rate is twice that for those who do get into hospital.",
  "f", "fat.tcoeff", "0.5", "Time co-efficient for increasing mortality rate as time in the **H** compartment increases for each individual in it. See `fat.rand` for details.",
  "", "vital", "TRUE", "Enables demographics, that is, arrivals and departures, to and from the simulated population.",
  "", "a.rate", "(10.5/365)/1000", "Background demographic arrival rate. Currently all arrivals go into the **S** compartment, the default is approximately the daily birth rate for Australia. Will be extended to cover immigration in future versions.",
  "", "ds.rate, de.rate, de.rate, dq.rate, dh.rate, dr.rate", "various rates", "Background demographic departure (death not due to virus) rates. Defaults based on Australian crude death rates. Can be used to model emigration as well as deaths.",
  "", "out", "mean", "Summary function for the simulation runs. median is also available, or percentiles, see the `EpiModel` documentation.",
                                       
  )
param_docs %>%
  gt() %>%
  fmt_markdown(columns = TRUE) %>%
  tab_options(table.width=pct(90))
```


